Circumnuclear Media and Accretion Rates of Quiescent Supermassive Black Holes

### 2015-04-28 ms.tex
Wording changes to the conclusions

  \begin{enumerate}
  \item A stagnation radius, $\rs$, divides the nuclear gas between an
    accretion flow and an outward wind. In steady-state the gas inflow
    rate towards the black hole is proportional to the stellar mass
    enclosed inside of $\rs$.  In the limit of strong heating, the
    stagnation radius resides interior to the SMBH influence radius
    and coincides with the Bondi radius (eq.~[\ref{eq:rbondi}]), and
    the Bernoulli parameter of the flow at large radii is positive
    (Appendix \ref{app:be}).  In the limit of weak heating ($\zeta <
    \zeta_c$; eq.~[\ref{eq:zetacrit}]), the stagnation radius moves to
    large radii, near or exceeding the stellar break radius $r_b \gg
    r_{\rm inf}$, greatly increasing the density of gas on smaller
    radial scales.

  \item Including the effects of heat transport by electron conduction
    results in at most order unity changes to the key properties of
    the flow (e.g. stagnation radius, mass accretion rate, and thermal
    stability) for causal values of the conduction saturation
    parameter $\phi < 0.1$ (eq.~[\ref{eq:conduction}]).  Angular
    momentum is also unlikely to invalidate our spherically symmetric
    model on large scales, given the inferred stellar rotation rates
    near the stagnation radius ($\S\ref{sec:rotation}$). 
%However, on
 %   small scales gas would encounter a centrifugal barrier which would
  %  cause much of the gas flowing inwards on large scales to be driven
   % outwards in an outflow (as in \citealt{Li+2013}).

@@

\item The location of the stagnation radius, and hence the SMBH
  accretion rate, is a sensitive function of the gas heating rate
  $\propto v_w^{2}$.  Quantitatively, $\rs\propto v_w^{-2}$, implying
  that $\Mdot\propto v_w^{-3.8 (-2.4)}$ for core (cusp) galaxies.
  However, $\Mdot$ can increase even more rapidly with decreasing
  $v_w$ for two reasons: (1) when $v_w$ becomes comparable to the
  stellar velocity dispersion ($\zeta < \zeta_c$;
  eq.~$[\ref{eq:zetacrit}]$), then gas remains bound to the stellar
  bulge and cannot produce a flow with a positive kinetic energy
  interior to the stellar break radius (Appendix \ref{app:be}); (2)
  For low $v_w$, gas near the stagnation radius becomes thermally
  unstable, which based on the results of previous numerical studies
  (e.g.,\citealt{Ciotti+10}) is instead likely to result in
  time-dependent cycles of alternating high and low accretion phases
  ($\S\ref{sec:cycle}$).

  \item Stellar wind heating, supernovae, and AGN feedback depend
    explicitly on the SMBH mass as well as the SFH in the galactic
    nuclear region.  A young starburst of age $\lesssim 10$ Myr
    produces mass and energy injection dominated by young stellar
    winds ($v_w \sim 1000$ km s$^{-1}$), while for exclusively old
    stellar populations mass input is dominated by AGB winds and
    heating is dominated by AGN feedback and supernovae.  Ongoing,
    continuous star formation presents a hybrid situation, with energy
    input usually dominated by young stars and mass input dominated by
    the old stars (Appendix \ref{app:windheat}).  Galactic nuclei with
    $\Mbh \lesssim 10^8\Msun$ heated according to their average SFHs
    (as derived from cosmological simulations) receive a stellar wind
    heating of $v_w \sim 700$ km s$^{-1}$.  This heating can be
    suppressed if the star formation is sufficiently intermittent
    (Fig.~\ref{fig:NickPlot2}).

\item Type Ia SNe only provide a continuous heating source exterior to
  the Ia radius $\rIa$ (eq.~[\ref{eq:rIa}]) where the time between
  subsequent SNe exceeds the dynamical timescale.  Non-Ia heating due
  to the average SFH results in $\rs <\rIa$, except for the most
  massive galaxies (Fig.~\ref{fig:vwSources}).  However, if $\rs$ does
  approach or exceed $\rIa$, then the SN Ia heating is usually
  sufficiently large to prevent $\rs$ from greatly exceeding $\rIa$.
  SNe Ia thus regulates the SMBH accretion rate below the value
  $\dot{M}_{\rm Ia}$ (eq.~[\ref{eq:eddr_Ia}]), except possibly in high
  mass elliptical galaxies with large break radii and $\zeta(v_w^{\rm
    Ia}) \gtrsim \zeta_c$.

@@

\item Stellar wind heating from the average SFH may be sufficient to
  permit thermally-stable, steady-state accretion, depending on the accretion efficiency of the SMBH.  However, the
  resulting average Eddington-normalized accretion rate is predicted
  to increase weakly with $M_{\bullet}$ (Fig.~\ref{fig:miller}), in
  tension with the (also weak) downsizing trend of measured nuclear
  X-ray luminosities of early-type galaxies (e.g.,
  \citealt{Miller+15}).  Thermally stable accretion models can reproduce the observed scatter
  in nuclear X-ray luminosities at fixed $\Mbh$ of 2-4 orders of
  magnitude due to combination of (1) intrinsic scatter in the SMBH
  scaling relations; (2) stochastic variation in SFH from galaxy to
  galaxy, or from time to time due to burstiness in the SFH
  (e.g.~Fig.~\ref{fig:NickPlot2}); (3) Variations in the amount of
  inflowing mass which is lost to small scale outflows from the SMBH
  accretion disk, due to variations in the angular momentum of the
  accreting gas (e.g.~\citealt{Li+13}).

\item However, for lower X-ray radiative efficiencies, the accretion rates of early-type galaxies are above the maximum value for thermally stable flow, $\dot{M}_{\rm TI}$
  (Fig.~\ref{fig:miller}).  This implies that current X-ray detections could instead be comprised
  mostly of nuclei undergoing outburst due to thermal instabilities.
  In such a case, there may exist a separate (usually undetected) branch of low-$L_X$ nuclei accreting stably.  The existence of such a bimodal population might call into question
  methods of extrapolating the $L_{\rm X}(M_{\bullet}$) relationship
  to low $M_{\bullet}$ to constrain the SMBH occupation
  fraction of low mass galaxies (\citealt{Miller+15}).

\item Low mass black holes grow in the low redshift universe over
  timescales comparable to the Hubble time (\citealt{Heckman+04}).  This
  accretion rates so required are too high to be consistent with the
  value predicted for the average SFH history and the maximum allowed
  for steady-state, thermally stable accretion
  (Fig.~\ref{fig:bh_growth}).  Perhaps unsurprisingly, low-$M_{\bullet}$ growth and AGN
  activity must instead be driven by a supply of gas external to the
  nucleus, e.g. due to galaxy mergers or thermal instabilities on
  larger, galactic scales (e.g.~\citealt{Voit+15}).

@@

We conclude by drawing attention to a few limitations of our
model.  First, we have neglected any large scale inflow of gas to the
nucleus by assuming the only source of gas feeding the black hole is
stellar wind mass loss from the local stellar population.  Our
estimates represent a lower bound on the time averaged gas density of
the CNM.  Our model cannot account for AGN.

Predictions for the SMBH accretion rate are strongly tied to the age
and radial distribution of the stellar population within the sphere of
influence.  Our model therefore cannot make definitive predictions for
the accretion mode of individual galaxies without knowledge of their
specific SFH.  We adopt halo average star formation rates measured in
cosmological simulations by \citet{MosterNaab+:2013a}, and neglect any
spatial variation in the star formation rate for any given
galaxy. However, galaxies of mass $\gsim 7\times 10^{10} \Msun$
assemble their stars from the inside out
\citep{PerezCidFernandes+:2013a}, resulting in galactic nuclei which
may be systematically older than assumed in our model and thus more
prone to thermal instability.

>>>>

  \begin{enumerate}
  \item A stagnation radius, $\rs$, divides the nuclear gas between an
    accretion flow and an outgoing wind. In steady-state the gas inflow
    rate towards the black hole is proportional to the stellar mass
    enclosed inside of $\rs$.  In the limit of strong heating, the
    stagnation radius resides interior to the SMBH influence radius
    and coincides with the Bondi radius (eq.~[\ref{eq:rbondi}]), and
    the Bernoulli parameter of the flow at large radii is positive
    (Appendix \ref{app:be}).  In the limit of weak heating ($\zeta <
    \zeta_c$; eq.~[\ref{eq:zetacrit}]), the stagnation radius moves to
    large radii, near or exceeding the stellar break radius $r_b \gg
    r_{\rm inf}$, greatly increasing the density of gas on smaller
    radial scales.

  \item Including the effects of heat transport by electron conduction
    results in at most order unity changes to the key properties of
    the flow (e.g. stagnation radius, mass accretion rate, and thermal
    stability) for causal values of the conduction saturation
    parameter $\phi < 0.1$ (eq.~[\ref{eq:conduction}]).  Angular
    momentum is also unlikely to invalidate our spherically symmetric
    model on large scales, given observationally inferred rotation rates of galactic nuclei
    near the stagnation radius ($\S\ref{sec:rotation}$). 
%However, on
 %   small scales gas would encounter a centrifugal barrier which would
  %  cause much of the gas flowing inwards on large scales to be driven
   % outwards in an outflow (as in \citealt{Li+2013}).

@@

\item The location of the stagnation radius, and hence the SMBH
  accretion rate, is a sensitive function of the gas heating rate
  $\propto v_w^{2}$.  Quantitatively, $\rs\propto v_w^{-2}$, implying
  that $\Mdot\propto v_w^{-3.8 (-2.4)}$ for fiducial core (cusp) galaxies.
  However, $\Mdot$ can increase even more rapidly with decreasing
  $v_w$ for two reasons: (1) when $v_w$ becomes comparable to the
  stellar velocity dispersion ($\zeta < \zeta_c$;
  eq.~$[\ref{eq:zetacrit}]$), then gas remains bound to the stellar
  bulge and cannot produce a flow with a positive kinetic energy
  interior to the stellar break radius, as we show in Appendix \ref{app:be}; (2)
  For low $v_w$, gas near the stagnation radius becomes thermally
  unstable, which based on the results of previous numerical studies
  (e.g.,\citealt{Ciotti+10}) is instead likely to result in
  time-dependent cycles of alternating high and low accretion phases
  ($\S\ref{sec:cycle}$).

  \item Stellar wind heating, supernovae, and AGN feedback depend
    explicitly on the SMBH mass as well as the SFH in the galactic
    nucleus.  A young starburst of age $\lesssim 10$ Myr
    produces mass and energy injection dominated by young stellar
    winds ($v_w \sim 1000$ km s$^{-1}$), while for exclusively old
    stellar populations mass input is dominated by AGB winds and
    heating is dominated by AGN feedback and supernovae.  Ongoing,
    continuous star formation presents a hybrid situation, with energy
    input usually dominated by young stars and mass input dominated by
    old stars (Appendix \ref{app:windheat}).  Galactic nuclei with
    $\Mbh \lesssim 10^8\Msun$ that are heated according to their average SFHs
    (as derived from cosmological simulations) receive a stellar wind
    heating of $v_w \sim 700$ km s$^{-1}$.  This heating can be
    suppressed if the star formation is sufficiently intermittent
    (Fig.~\ref{fig:NickPlot2}).

\item Type Ia SNe only provide a continuous heating source exterior to
  the Ia radius $\rIa$ (eq.~[\ref{eq:rIa}]) where the time between
  subsequent SNe exceeds the dynamical timescale.  Non-Ia heating due
  to the average SFH results in $\rs <\rIa$, except for the most
  massive galaxies (Fig.~\ref{fig:vwSources}).  However, if $\rs$ does
  approach $\rIa$, then SN Ia heating is usually
  large enough to prevent $\rs$ from greatly exceeding $\rIa$.
  SNe Ia thus regulate the SMBH accretion rate below the value
  $\dot{M}_{\rm Ia}$ (eq.~[\ref{eq:eddr_Ia}]), except possibly in high
  mass elliptical galaxies with large break radii and $\zeta(v_w^{\rm
    Ia}) \gtrsim \zeta_c$.

@@

\item Stellar wind heating from the average SFH may be sufficient to
  permit thermally-stable, steady-state accretion, depending on the accretion efficiency of the SMBH.  However, the
  resulting average Eddington-normalized accretion rate is predicted
  to increase weakly with $M_{\bullet}$ (Fig.~\ref{fig:miller}), in
  tension with the (also weak) downsizing trend of measured 
  X-ray luminosities in early-type galactic nuclei (e.g.,
  \citealt{Miller+15}).  Thermally stable accretion models can reproduce the observed scatter
  in nuclear X-ray luminosities at fixed $\Mbh$ (2-4 orders of
  magnitude) due to the combination of (1) intrinsic scatter in the SMBH
  scaling relations; (2) stochastic variation in SFH between galaxies, or within one galaxy due to burstiness in the SFH, as in ~Fig.~\ref{fig:NickPlot2}; (3) Variations in the amount of
  inflowing mass which is lost to small scale outflows from the SMBH
  accretion disk, likely due to variations in the angular momentum of the
  accreting gas (e.g.~\citealt{Li+13}).

\item However, for lower X-ray radiative efficiencies, the accretion rates of early-type galaxies are above the maximum value for thermally stable flow, $\dot{M}_{\rm TI}$
  (Fig.~\ref{fig:miller}).  This implies that current X-ray detections could instead be comprised
  mostly of nuclei undergoing outburst due to thermal instabilities.
  In such a case, there may exist a separate (usually undetected) branch of low-$L_X$ nuclei accreting stably.  The existence of such a bimodal population would call into question constraints on the SMBH occupation
  fraction in low mass galaxies (\citealt{Miller+15}) derived by
  extrapolating the $L_{\rm X}(M_{\bullet}$) relationship
  to small $M_{\bullet}$.

\item Low mass black holes grow in the low redshift universe over
  timescales comparable to the Hubble time (\citealt{Heckman+04}).  The
  accretion rates so required are too high to be consistent with either the
  value predicted for the average SFH history or the maximum allowed
  for steady-state, thermally stable accretion
  (Fig.~\ref{fig:bh_growth}).  Perhaps unsurprisingly, low-$M_{\bullet}$ growth and AGN
  activity must instead be driven by a supply of gas external to the
  nucleus, such as galaxy mergers or thermal instabilities on
  larger, galactic scales (e.g.~\citealt{Voit+15}).

@@

We conclude by drawing attention to a few limitations of our
model.  First, we have neglected any large scale inflow of gas to the
nucleus by assuming the only source of gas feeding the black hole is
stellar wind mass loss from the local stellar population.  Our
estimates represent a lower bound on the time averaged gas density of
the CNM.  Although our model can, under certain assumptions, describe quiescent galactic nuclei, it cannot account for AGN.

Predictions for the SMBH accretion rate are strongly tied to the age
and radial distribution of the stellar population within the sphere of
influence.  Our model therefore cannot make definitive predictions for
the accretion mode of individual galaxies without knowledge of their
specific SFH.  We adopt halo-averaged star formation rates measured in
cosmological simulations by \citet{MosterNaab+:2013a}, and neglect any
spatial variation in the star formation rate for any given
galaxy. However, galaxies of mass $\gsim 7\times 10^{10} \Msun$
assemble their stars from the inside out
\citep{PerezCidFernandes+:2013a}, resulting in galactic nuclei which
may be systematically older than assumed in our model and thus more
prone to thermal instability.
